Covid-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)

### 2020-10-14 manuscript/manuscript.Rmd
add index of multiple deprivation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
options(OutDec = "·")
knit_output_pdf <- F
if(knit_output_pdf){
  options(knitr.table.format = "pdf")
} else {
  options(knitr.table.format = "pandoc")
}

@@

homes_country <- select(reference_homes, home_code, postcode) %>% 
  left_join(transmute(reference_geography, postcode, 
                      country, region_england = region), by = "postcode") %>% 
  mutate(region_UK = if_else(country == "England", region_england, country)) %>% 
  select(-postcode)
reference_homes <- left_join(reference_homes, homes_country)
rm(homes_country)

@@

reference_homes <- reference_homes %>% 
  mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
         occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)),
         region_UK = as.factor(region_UK)) %>% 
  mutate(occ_bedroom_ratio_g6 = relevel(occ_bedroom_ratio_g6, '(0·7,0·85]'),
         occ_bedroom_ratio_g5 = relevel(occ_bedroom_ratio_g5, '(0·8,1]'),
         region_UK <- relevel(region_UK, "Scotland"))

residents_overview <- tline %>% 
  filter(rday==1) %>%
  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  group_by(resident_id, home_code, ageg, gender, 
           admission_type2, home_type, user_type2, status) %>% 
  summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed),
            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
            confirmed = max(-susceptible_confirmed+1, na.rm = T),
            start_date = min(date, na.rm = T),
            stop_date = max(date, na.rm = T),
            start_int = as.numeric(min(date, na.rm = T) - date_min),
            stop_int = as.numeric(max(date, na.rm = T)- date_min)
    ) %>% 
  ungroup() %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
                   beds,  pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))

@@

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  ))  %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date+1),
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric( entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed")),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
                   beds, pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% 
  ungroup()

@@

res_counts <- residents_overview %>% 
  group_by(home_code) %>% 
  summarise(unique_residents = n_distinct(resident_id))

@@

residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
  left_join(select(new_cases, date, home_code, 
                     new_confirmed_home, new_confirmed_hospital,
                     new_symptomatic_home, new_symptomatic_hospital,
                     new_deaths_home, new_deaths_hospital)) %>% 
  filter(!is.na(home_code)) %>% 
  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, 
                     staff_total, nursing_dementia, nursing_general, 
                     residential_dementia, residential_general, younger_person)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  left_join(occupancy_interpolated) %>% 
  arrange(home_code, date) %>% 
  group_by(home_code) %>% 
  mutate(new_symptomatic = tidyr::replace_na(new_symptomatic_home, 0) + 
           tidyr::replace_na(new_symptomatic_hospital, 0),
         new_confirmed = tidyr::replace_na(new_confirmed_home, 0) + 
           tidyr::replace_na(new_confirmed_hospital, 0),
         new_deaths = tidyr::replace_na(new_deaths_home, 0)  +
           tidyr::replace_na(new_deaths_hospital, 0)
         ) %>% 
  mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
         I_confirmed = total_cases_in_home(occupancy, new_confirmed, date)) %>% 
  ungroup()

@@

`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]) residents. `r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1·9 [1·7; 2·2]). 33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.

@@

Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]

@@

Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.  `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`). `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in these residents, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]. The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22·6% versus 11·2%).

@@

arsenal::tableby(
  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Funding type` +  
    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
  data = mutate(residents_overview,
                Sex = gender,
                Age = ageg,
                `Resident type` = user_type2,
                `Admission type` = admission_type2,
                `Funding type` = home_type,
                `Infection status by 14 June` = covid_status, 
                `Status as of 14 June` = status,
                `Region/nation` = region_UK 
  )) %>% summary()
```

@@

Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model. Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). Large care homes had greater rates of infection (adjusted HR = 1.8 [1·4; 2·4] for care homes with $\geq$ 70 beds versus <35 beds). Care home baseline occupancy and staffing ratios had the greatest effect on residents risk of infection. For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1·9; 3·3] greater in homes with 0·85-1 resident per room versus homes with 0·7-0·85 resident per room. In homes with more residents than bedrooms, hazards were 10 times [6·9; 15] greater.

Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1·23 [1·17; 1·31]).

@@

```{r hazard_ratios_confirmed,   results="asis"}
# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+
#                     home_type+user_type2+  bed_tot_category + bed_to_staff_ratio +occ_bedroom_ratio_g6, data = residents_overview)
#  summary(cph_conf)
# cox.zph(cph_conf)
# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
explanatory = c("gender", "ageg", "home_type", "user_type2",  
                "bed_tot_category", "bed_to_staff_ratio",  "occ_bedroom_ratio_g6")
dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
CPH_confirmed <-  finalfit(residents_overview,
                           dependent, 
                           explanatory, column = T,
                           add_dependent_label = F)  %>%     ff_remove_p() 
knitr::kable(CPH_confirmed,
             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality. After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·4 [1·3; 1·6]), and receiving nursing care (adjusted HR = 1·4 [1·2; 1.5]).

We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection (group A) in non-outbreak homes as the reference group. Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2·0 [1·7; 2·2]). Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4·3 [3·0; 6·2] versus 9·4 [7·6; 12]).  All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3·3 [2·0; 5·7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.

@@

```{r hazard_ratios_death, results="asis"}
# cph_death2 <- coxph(Surv(date1, date2, death_status) ~ gender+ ageg + home_type + user_type2 + 
#                       bed_tot_category + bed_to_staff_ratio + occ_bedroom_ratio_g6 + 
#                       covid_status_outbreak  + cluster(resident_id), data = residents_episodes)
# summary(cph_death2)
# cox.zph(cph_death2)
residents_episodes$bed_to_staff_ratio <- 1/residents_episodes$staff_to_bed_ratio
explanatory = c("gender", "ageg", "home_type", "user_type2",
                "bed_tot_category", "bed_to_staff_ratio", "occ_bedroom_ratio_g6",
                "covid_status_outbreak",  "cluster(resident_id)")
dependent = "Surv(time = date1, time2 = date2, event = death_status)"
CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
                       column = T, add_dependent_label = F)  %>%   ff_remove_p() 
knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

Model-based estimates of attributable mortality were derived from the individual-level data. Overall, 567/1,694 (33%) deaths were attributed to Covid-19. In homes with outbreaks only 28% (159 residents) of the mortality attributable to COVID-19 occurred in people with confirmed infection (Groups C and D), (`r table_nums("tab_attributable_deaths", display = "cite")`). Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41·7%). Model-based estimates of deaths based on individual-level were slightly higher (8%) than counts from the aggregate data.

@@

HRs_covid <- tibble::enframe(cph_death3$coefficients[grep("covid_status_outbreak", 
                                                          names(cph_death3$coefficients))]) %>% 
  transmute(covid_status_outbreak_digit = gsub("covid_status_outbreak_digit", "", name),
            HR = exp(value))
HRs_CI_covid <- as_tibble(exp(confint(cph_death3))[grep("covid_status_outbreak", 
                                                          names(cph_death3$coefficients)),]) %>% 
  rename(HR_lower =  `2·5 %`,
         HR_upper = `97·5 %`)
HRs_CI_covid$covid_status_outbreak_digit <- gsub(
  "covid_status_outbreak_digit", "",
  grep("covid_status_outbreak", names(cph_death3$coefficients), value = T)
) 
HRs_covid <- merge(HRs_covid, HRs_CI_covid)

@@

This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35·7%. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks, implying substantial case under-ascertainment. Less than one-third of deaths attributable to COVID-19 in outbreak homes were confirmed which is likely due to poor availability of testing until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy are critically important to reducing the spread of infection.

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
knit_output_pdf <- F
if(knit_output_pdf){
  options(knitr.table.format = "pdf")
} else {
  options(knitr.table.format = "pandoc")
}

@@

homes_country <- select(reference_homes, home_code, postcode) %>% 
  left_join(transmute(reference_geography, postcode, 
                      country, region_england = region,
                      uk_imd_england_quintile), by = "postcode") %>% 
  mutate(region_UK = if_else(country == "England", region_england, country)) %>% 
  select(-postcode)
reference_homes <- left_join(reference_homes, homes_country)
rm(homes_country)

@@

reference_homes <- reference_homes %>% 
  mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
         occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)),
         region_UK = as.factor(region_UK)) %>% 
  mutate(occ_bedroom_ratio_g6 = relevel(occ_bedroom_ratio_g6, '(0.7,0.85]'),
         occ_bedroom_ratio_g5 = relevel(occ_bedroom_ratio_g5, '(0.8,1]'),
         region_UK <- relevel(region_UK, "Scotland"))

residents_overview <- tline %>% 
  filter(rday==1) %>%
  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  group_by(resident_id, home_code, ageg, gender, 
           admission_type2, home_type, user_type2, status) %>% 
  summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed),
            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
            confirmed = max(-susceptible_confirmed+1, na.rm = T),
            start_date = min(date, na.rm = T),
            stop_date = max(date, na.rm = T),
            start_int = as.numeric(min(date, na.rm = T) - date_min),
            stop_int = as.numeric(max(date, na.rm = T)- date_min)
    ) %>% 
  ungroup() %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
                   beds,  pct_occupancy, no_bedrooms, uk_imd_england_quintile,
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))

@@

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  ))  %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date+1),
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric( entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed")),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
                   beds, pct_occupancy, no_bedrooms, uk_imd_england_quintile,
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% 
  ungroup()

@@

res_counts <- residents_overview %>% 
  group_by(home_code) %>% 
  summarise(unique_residents = n_distinct(resident_id),
            total_rdays = sum(resident_days))

@@

residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
  left_join(select(new_cases, date, home_code, 
                     new_confirmed_home, new_confirmed_hospital,
                     new_symptomatic_home, new_symptomatic_hospital,
                     new_deaths_home, new_deaths_hospital)) %>% 
  filter(!is.na(home_code)) %>% 
  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, 
                     staff_total, nursing_dementia, nursing_general, 
                     residential_dementia, residential_general, younger_person)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  left_join(occupancy_interpolated) %>% 
  mutate(new_symptomatic = tidyr::replace_na(new_symptomatic_home, 0) + 
           tidyr::replace_na(new_symptomatic_hospital, 0),
         new_confirmed = tidyr::replace_na(new_confirmed_home, 0) + 
           tidyr::replace_na(new_confirmed_hospital, 0),
         new_deaths = tidyr::replace_na(new_deaths_home, 0)  +
           tidyr::replace_na(new_deaths_hospital, 0)
         ) %>%   
  arrange(home_code, date) %>% 
  group_by(home_code) %>% 
  mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
         I_confirmed = total_cases_in_home(occupancy, new_confirmed, date)) %>% 
  ungroup()

@@

`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]) residents. `r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1.9 [1.7; 2.2]). 33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.

@@

Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]

@@

Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.  `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`). `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in these residents, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]. The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22.6% versus 11.2%).

@@

arsenal::tableby(
  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Funding type` +  
    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation` + `Index of Multiple Deprivation`, 
  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
  data = mutate(residents_overview,
                Sex = gender,
                Age = ageg,
                `Resident type` = user_type2,
                `Admission type` = admission_type2,
                `Funding type` = home_type,
                `Infection status by 14 June` = covid_status, 
                `Status as of 14 June` = status,
                `Region/nation` = region_UK,
                `Index of Multiple Deprivation` = uk_imd_england_quintile
  )) %>% summary()
```

@@

Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model. Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1.5 [1.2; 1.8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). Large care homes had greater rates of infection (adjusted HR = 1.8 [1.4; 2.4] for care homes with $\geq$ 70 beds versus <35 beds). Care home baseline occupancy and staffing ratios had the greatest effect on residents risk of infection. For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1.9; 3.3] greater in homes with 0.85-1 resident per room versus homes with 0.7-0.85 resident per room. In homes with more residents than bedrooms, hazards were 10 times [6.9; 15] greater.

Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1.23 [1.17; 1.31]).

@@

```{r hazard_ratios_confirmed,   results="asis"}
# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+
#                     home_type+user_type2+  bed_tot_category + bed_to_staff_ratio +occ_bedroom_ratio_g6, data = residents_overview)
#  summary(cph_conf)
# cox.zph(cph_conf)
# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
residents_overview$uk_imd_england_quintile <- factor(residents_overview$uk_imd_england_quintile,
                                                     levels = c(
                                                       '3', 
                                                       '1 - least deprived', '2', '4', '5 - most deprived'
                                                     ))
residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
explanatory = c("gender", "ageg", "home_type", "user_type2",  "uk_imd_england_quintile",
                "bed_tot_category", "bed_to_staff_ratio",  "occ_bedroom_ratio_g6")
dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
CPH_confirmed <-  finalfit(residents_overview,
                           dependent, 
                           explanatory, column = T,
                           add_dependent_label = F)  %>%     ff_remove_p() 
knitr::kable(CPH_confirmed,
             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality. After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.4 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.2; 1.5]).

We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection (group A) in non-outbreak homes as the reference group. Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2.0 [1.7; 2.2]). Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4.3 [3.0; 6.2] versus 9.4 [7.6; 12]).  All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3.3 [2.0; 5.7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.

@@

```{r hazard_ratios_death, results="asis"}
# cph_death2 <- coxph(Surv(date1, date2, death_status) ~ gender+ ageg + home_type + user_type2 +
#                       bed_tot_category + bed_to_staff_ratio + occ_bedroom_ratio_g6 +
#                       covid_status_outbreak  + cluster(resident_id), data = residents_episodes)
# summary(cph_death2)
# cox.zph(cph_death2)
residents_episodes$uk_imd_england_quintile <- factor(residents_episodes$uk_imd_england_quintile,
                                                     levels = c(
                                                       '3', 
                                                       '1 - least deprived', '2', '4', '5 - most deprived'
                                                     ))

residents_episodes$bed_to_staff_ratio <- 1/residents_episodes$staff_to_bed_ratio
explanatory = c("gender", "ageg", "home_type", "user_type2", "uk_imd_england_quintile",
                "bed_tot_category", "bed_to_staff_ratio", "occ_bedroom_ratio_g6",
                "covid_status_outbreak",  "cluster(resident_id)")
dependent = "Surv(time = date1, time2 = date2, event = death_status)"
CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
                       column = T, add_dependent_label = F)  %>%   ff_remove_p() 
knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

Model-based estimates of attributable mortality were derived from the individual-level data. Overall, 567/1,694 (33%) deaths were attributed to Covid-19. In homes with outbreaks only 28% (159 residents) of the mortality attributable to COVID-19 occurred in people with confirmed infection (Groups C and D), (`r table_nums("tab_attributable_deaths", display = "cite")`). Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41.7%). Model-based estimates of deaths based on individual-level were slightly higher (8%) than counts from the aggregate data.

@@

HRs_covid <- tibble::enframe(cph_death3$coefficients[grep("covid_status_outbreak", 
                                                          names(cph_death3$coefficients))]) %>% 
  transmute(covid_status_outbreak_digit = gsub("covid_status_outbreak_digit", "", name),
            HR = exp(value))
HRs_CI_covid <- as_tibble(exp(confint(cph_death3))[grep("covid_status_outbreak", 
                                                          names(cph_death3$coefficients)),]) %>% 
  rename(HR_lower =  `2.5 %`,
         HR_upper = `97.5 %`)
HRs_CI_covid$covid_status_outbreak_digit <- gsub(
  "covid_status_outbreak_digit", "",
  grep("covid_status_outbreak", names(cph_death3$coefficients), value = T)
) 
HRs_covid <- merge(HRs_covid, HRs_CI_covid)

@@

This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35.7%. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks, implying substantial case under-ascertainment. Less than one-third of deaths attributable to COVID-19 in outbreak homes were confirmed which is likely due to poor availability of testing until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy are critically important to reducing the spread of infection.

@@

## Supplementary analysis

```{r effect_organisational_variables}
# adhoc1 <- residents_ct %>% 
#   filter(home_code %in% home_inclusion) %>% 
#   group_by(home_code) %>% 
#   summarise(confirmed_cases = sum(new_confirmed),
#             symptomatic_cases = sum(new_symptomatic),
#             denominator = sum(occupancy - I_confirmed)) %>% 
#   left_join(reference_homes) %>% 
#   left_join(select(filter(beds, year=="2020"), -year)) %>% 
#   mutate(staff_to_bed_ratio = staff_total/beds,
#          carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
#          bed_tot_category = cut(beds, 
#            breaks = seq(20,100,25), right = F, 
#            paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
#          ) %>% 
#   mutate(staff_to_bed_ratio_g5 = cut(
#     staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
#     quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
#     max(staff_to_bed_ratio)))) %>% 
#   left_join(home_mean_occupancy) %>% 
#   mutate(pct_occupancy = mean_occupancy/beds,
#          occ_bedroom_ratio = mean_occupancy/no_bedrooms) %>% 
#   mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
#          occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)))
# 
# ggplot(adhoc1, aes(staff_to_bed_ratio, confirmed_cases/denominator*100000)) +
#   geom_point() +
#   geom_smooth() +
#   ylab("Cumulative incidence per 100,000") +
#   xlab("Ratio staff : beds")
# 
# mod_staffing <- glm(confirmed_cases ~  
#   staff_to_bed_ratio + offset(log(denominator)), 
#   family = poisson(link = "log"), data = adhoc1)
# 
# 
# ggplot(adhoc1, aes(occ_bedroom_ratio, confirmed_cases/denominator*100000)) +
#   geom_point() +
#   geom_smooth() +
#   ylab("Cumulative incidence per 100,000") +
#   xlab("Ratio occupants : bedrooms (Jan-Mar 2020)")
# 
# adhoc1 %>% 
#   group_by(occ_bedroom_ratio_g5) %>% 
#   summarise(confirmed_cases = sum(confirmed_cases),
#             denominator = sum(denominator))  %>% 
#   mutate(rate = confirmed_cases/denominator*100000)
